---
title: "p8105_hw1_al4922.Rmd"
output: github_document
date: "2026-09-23"
---

```{r include=FALSE}
library(tidyverse)
```
## Problem 1

```{r}
data("penguins", package = "palmerpenguins") #load penguins dataset
summary(penguins) #gives species and mean flipper length
nrow(penguins) #gives number of penguins
ncol(penguins) #gives number of variables 
```

## Penguins Dataset Description 

The penguins dataset describes the various characteristics (habitat and various physical characteristics) of three species of penguins Adelie, Chinstrap and Gentoo. Its key variables are: species, island, bill lengths and depth, body mass, flipper length, year and sex. The data includes `r nrow(penguins)` individuals and `r ncol(penguins)` distinct variables and the mean flipper length of the penguins is `r round(mean(penguins$flipper_length_mm, na.rm = TRUE), 2)` mm.


```{r flippvervbill_scatterplot}
penguin_scatterplot <- ggplot(data = penguins, 
       mapping = aes(x = bill_length_mm, 
                      y = flipper_length_mm, 
                      color = species)) + 
  geom_point()

penguin_scatterplot
```
```{r export_to_directory}
ggsave("flippervbill.png",
       plot = penguin_scatterplot) #exports plot to directory
```
## Problem 2

**Step 1: Create Dataframe**

```{r set_seed}
set.seed(367) 
#This makes it so that the numbers generated are reproducible
```
```{r norm_sample}
norm_sample <- rnorm(10, mean = 0, sd = 1)
#This generates a group with 10 samples form a standard Normal distribution and assigns it to the new variable of norm_sample
norm_sample
```

```{r dataframe_creation}
q2_df <- data.frame(
  norm_vec = norm_sample, #since data.frame() looks for variable norm_sample when it runs all arguments simultaneously that's why we had to create norm_sample in the last step
  log_vec = norm_sample > 0, #storing logical vector     that asks if elements of sample greater than zero?
  char_vec = letters[1:10], #generates character vector of length 10
  factor_vec = factor(c("low", "med", "high", "low", "med", "high", "low", "med", "high", "low"))
) #generates factor vector of length 10, with 3 different levels
#This generates the data frame 
```

**Step 2: Assess Mean of Each Column**
```{r assess_mean}
# 1. Numeric variable (Works)
q2_df %>% 
  pull(norm_vec) %>% 
  mean()

# 2. Logical variable (Works - returns proportion of TRUEs)
q2_df %>% 
  pull(log_vec) %>% 
  mean()

# 3. Character variable (Returns NA with warning)
q2_df %>% 
  pull(char_vec) %>% 
  mean()

# 4. Factor variable (Returns NA with warning)
q2_df %>% 
  pull(factor_vec) %>% 
  mean()
```
**Step 3: Convert To Numeric**
```{r convert_numeric, eval = FALSE}
# Convert logical to numeric
as.numeric(q2_df$log_vec)

# Convert character to numeric
as.numeric(q2_df$char_vec)

# Convert factor to numeric
as.numeric(q2_df$factor_vec)
```
## What Happens and Why
1. Logical ('log_vec'):
The logical vectors were successfully converted into '1's and '0's because in R, boolean is stored in binary representation and so translating from 'TRUE'/'FALSE' to '1'/'0' works. 

2. Character ('char_vec'):
Each string was converted into NA because there's no "quantity" assigned to different 'words' or 'strings'. R can't do this conversion. 

3. Factor ('factor_vec'):
The factor levels were converted into three corresponding numbers (ie. '1', '2', '3') because R stores factor variables as integer vectors in the background. What 'as.numeric()' does is that it takes the text labels 'off' exposing the numeric codes. 

## Does this explain what happens when we try to take means? 
1. **Logical variables work directly with `mean()`** because mean() performs implicit coercion. This means that in the background, it converts `TRUE` -> `1` and `FALSE` -> `0`. With `TRUE` and `FALSE` converted into numbers like with the as.numeric function, a mean can be calculated. 

2. **Character variables fail with `mean()`** because implicit coercion encounters non-numeric text and generates `NA`s, triggering the error/warning. This means that it cannot be 'forced' into a number and so no mean can be calculated. 

3. **Factor variables reveal why R has protective guards.** While `as.numeric()` can turn a factor into numbers (1, 2, 3), `mean()` doesn't perform implicit coercion here. Here, `mean()` checks `is.numeric()` first, sees `FALSE`, and returns `NA`. This prevents you from accidentally calculating a mathematically meaningless "average" on categorical codes (like averaging "Low", "Medium", and "High"). However, if you explicitly force it using a nested function like `mean(as.numeric(factor_vec))`, R *will* calculate the average of those underlying level numbers created by the as.numeric function. 


